### 2025-09-26 YEAR2/MAT292_LectureNotes/main.tex
sept 26

>>>>

\begin{definition}[Computing Eigenvalues via Trace and Determinant]
    Consider matrix $A \in \mathbb{R}^{2 \times 2}$. We compute the eigenvalues by solving the characteristic polynomial:
    \begin{align*}
    0 &= \det(A - \lambda I) \\
    &= \lambda^2 - \text{tr}(A)\lambda + \det(A) \\
    \implies \lambda_{1,2} &= \frac{\text{tr}(A) \pm \sqrt{\text{tr}(A)^2 - 4\det(A)}}{2}
    \end{align*}

    If the roots are real, there are four cases that coud emerge from this (WLOG): 
    \begin{itemize}
        \item $\lambda_1 > \lambda_2 > 0$: We see much faster growth in the span assocaited with $\lambda_1$.
        \item $\lambda_1 < \lambda_2 < 0$: We see much faster decay in the span assocaited with $\lambda_1$.
        \item $\lambda_1 < 0 < \lambda_2$: We would see the span assocaited with $\lambda_2$ dominate as $t \to \infty$ and the span assocaited with $\lambda_1$ dominate as $t \to -\infty$. This is a saddle node.
        \item $\lambda_2  = 0 < \lambda_1$: Along the span of $v_2$, we would see horizontal line, as we growth only along the span of $v_1$. 
        \item $\lambda_1 = \lambda_2 > 0$: Along the span of $v_2$, we would see horizontal line, as we decay only along the span of $v_1$.
        \item $\lambda_1 = \lambda_2 = 0$: The system is trivial, and all points are fixed points.
        \item $\lambda_1 = \lambda_2 < 0$: It is a inward star that everything goes to the origin.
        \item $\lambda_1 = \lambda_2 > 0$: It is an outward star that everything goes away from the origin.
    \end{itemize}
\end{definition}

\begin{definition}[Complex Eigenvalues]
    Consider the case where the eigenvalues are complex, i.e., \( \lambda = \alpha \pm i\beta \) with \( \beta \neq 0 \). The general solution to the system can be expressed as:
    $$\textbf{x}(t) = e^{\alpha t} \left( c_1 \text{Re}(\vec{v} e^{i\beta t}) + c_2 \text{Im}(\vec{v} e^{i\beta t}) \right)$$
    where \( \vec{v} \) is the complex eigenvector corresponding to \( \lambda \), and \( c_1, c_2 \) are constants determined by the initial conditions.

    This would lead to the following definition of a radial and angular component:
    \begin{equation}
    v(t) = ve^{\alpha t} \mathrm{cis} (\beta t)
    \end{equation}
    where \( r(t) \) represents the radial distance from the origin and \( \theta(t) \) represents the angular position. So, for a general solution, we have:
    \begin{equation}
    \textbf{x}(t) = \sum_i c_i v_i e^{\alpha_i t} \mathrm{cis} (\beta_i t)
    \end{equation}
    The behavior of the system depends on the sign of \( \alpha \) and $\beta$:
    \begin{itemize}
        \item If \( \alpha > 0 \) the system spirals outward (complex source).
        \begin{itemize}
            \item If \( \beta > 0 \), the systems sprial counter-clockwise.
            \item If \( \beta < 0 \), the systems sprial clockwise.
        \end{itemize}
        \item If \( \alpha < 0 \) the system spirals inward (complex sink).
        \begin{itemize}
            \item If \( \beta > 0 \), the systems sprial counter-clockwise.
            \item If \( \beta < 0 \), the systems sprial clockwise.
        \end{itemize}
        \item If \( \alpha = 0 \) the system is a center (closed orbits).
        \begin{itemize}
            \item If \( \beta > 0 \), the systems sprial counter-clockwise.
            \item If \( \beta < 0 \), the systems sprial clockwise.       
        \end{itemize}
    \end{itemize}
\end{definition}

\paragraph{Complex Eigenvalues Resolves to Real Numbers} Consider eigenvalues $\lambda = \alpha \pm i\beta \in \mathbb{C}$ with $v = \textbf{p} + i\textbf{q}\in \mathbb{C}^n$. Then, the general solution $x: \mathbb{R} \to \mathbb{R}^n$ is given by:
\begin{equation}
\textbf{x}(t) = c_1 e^{\lambda t} \textbf{v} + c_2 e^{\bar{\lambda} t} \bar{\textbf{v}}
\end{equation}
where $c_1, c_2 \in \mathbb{C}$.

We also have:
$$
    \bar{\textbf{x}} = \bar{c_1} e^{\bar{\lambda} t} \bar{\textbf{v}} + \bar{c_2} e^{\lambda t} \textbf{v}
$$
So, we can say that:
$$
c_1 = \bar{c_2}, \quad c_2 = \bar{c_1}
$$
Let $c_1 =  \delta + i\gamma$, then $c_2 = \delta - i\gamma$. Thus, we have:
\begin{equation}
\textbf{x}(t) = (\delta + i\gamma)e^{(\alpha + i\beta)t}(\textbf{p} + i\textbf{q}) + (\delta - i\gamma)e^{(\alpha - i\beta)t}(\textbf{p} - i\textbf{q}) \\
\end{equation}
Seperating the real and imaginary parts, we have:
\begin{equation}
\textbf{x}(t) = \delta (ve^{\lambda t} + \bar{v}e^{\bar{\lambda} t}) + i\gamma (ve^{\lambda t} - \bar{v}e^{\bar{\lambda} t})
\end{equation}
By the Euler's formula, we have:
\begin{equation}
\textbf{x}(t) = 2\delta \mathrm{Re}(ve^{\lambda t}) - 2\gamma \mathrm{Im}(ve^{\lambda t}) \in \mathbb{R}^n
\end{equation}

\paragraph{Generalized eigenvectors and solution for a defective matrix}
Consider the matrix
\[
A=\begin{bmatrix}2 & 1\\[4pt]0 & 2\end{bmatrix},
\]
which has a double eigenvalue \(\lambda=2\) but only one linearly independent eigenvector
\[
v=\begin{bmatrix}1\\[4pt]0\end{bmatrix}.
\]

To complete a Jordan chain of length two we seek a generalized eigenvector \(v_g\) satisfying
\begin{equation}
(A-\lambda I)\,v_g = v. \label{eq:gen-eig}
\end{equation}
For this \(A\) we have \(A-2I=\begin{bmatrix}0&1\\0&0\end{bmatrix}\). Choosing the convenient solution
\[
v_g=\begin{bmatrix}0\\[4pt]1\end{bmatrix}
\]
indeed yields \((A-2I)v_g=v\), so \eqref{eq:gen-eig} holds.

The two linearly independent solutions corresponding to this Jordan chain are
\begin{equation}
x_1(t)=v\,e^{\lambda t} = \begin{bmatrix}1\\[4pt]0\end{bmatrix} e^{2t}, \label{eq:sol1}
\end{equation}
and
\begin{equation}
x_2(t)=(v_g + t\,v)\,e^{\lambda t} = \left(\begin{bmatrix}0\\[4pt]1\end{bmatrix} + t\begin{bmatrix}1\\[4pt]0\end{bmatrix}\right)e^{2t}. \label{eq:sol2}
\end{equation}
One checks directly that these satisfy \(x_i'(t)=Ax_i(t)\) using \(Av=\lambda v\) and \(Av_g=\lambda v_g + v\).

Therefore the general solution of \(\mathbf{x}'=A\mathbf{x}\) is the linear combination
\begin{equation}
\mathbf{x}(t)=C_1\,x_1(t) + C_2\,x_2(t)
= e^{\lambda t} \begin{bmatrix}
C_1 + C_2 t \\[4pt] C_2
\end{bmatrix}
\end{equation}
The phase portrait of this system is shown below:
% insert later

\begin{definition}[Trace-Determinant Plane]
    We found that we can find the eigenvalues of a \( 2 \times 2 \) matrix \( A \) using:
    $$
    \lambda_{1,2} = \frac{\text{tr}(A) \pm \sqrt{\text{tr}(A)^2 - 4\det(A)}}{2}
    $$
    where
    \begin{itemize}
        \item \( \text{tr}(A) = \lambda_1 + \lambda_2 = a_{11} + a_{22} \) is the trace of \( A \)
        \item \( \det(A) = \lambda_1 \lambda_2 = a_{11}a_{22} - a_{12}a_{21} \) is the determinant of \( A \)   
    \end{itemize}
    And we have the following analysis:
    \begin{itemize}
        \item If \( \det(A) < 0 \) and \( \text{tr}(A) > 0 \), then \( \lambda_1 > 0 > \lambda_2 \) (saddle node).
        \item Consider the discriminant boundry \( \text{tr}(A)^2 - 4\det(A) = 0 \), it is a parabola that upward opening with vertex at the origin. Consider region above that, where there is complex eigenvalues:
        \begin{itemize}
            \item If \( \text{tr}(A) > 0 \) and \( \det(A) > 0 \), then \( \alpha = \text{tr}(A)/2 > 0 \) (spiral source).
            \item If \( \text{tr}(A) < 0 \) and \( \det(A) > 0 \), then \( \alpha = \text{tr}(A)/2 < 0 \) (spiral sink).
            \item If \( \text{tr}(A) = 0 \) and \( \det(A) > 0 \), then \( \lambda_1 = -\lambda_2 = i\beta \) (center).
        \end{itemize}
        \item Consider the region below the discriminant boundry, where there is real eigenvalues:
        \begin{itemize}
            \item If \( \text{tr}(A) > 0 \) and \( \det(A) > 0 \), then \( \lambda_1 \geq \lambda_2 > 0 \) (nodal source).
            \item If \( \text{tr}(A) < 0 \) and \( \det(A) > 0 \), then \( 0 > \lambda_1 \geq \lambda_2 \) (nodal sink).
        \end{itemize}
        \item \textbf{Degenerate Star Node} Consider the line \( \text{tr}(A)^2 - 4\det(A) = 0 \).
        \item \textbf{Lines of Nodes} Consider the line \( \det(A) = 0 \).
    \end{itemize}
\end{definition}

### 2025-09-26 YEAR2/PHY293_LectureNotes/main.tex
sept 26

\section{The Wave Equation}

>>>>

\begin{example}
    Consider three equal masses \( m \) connected by four identical springs with spring constant \( k \) and fixed to walls on either side. The equations of motion for the three masses are given by:
    $$
        \begin{cases}
            m\ddot{x}_1 &= -2k x_1 + k x_2 \\
            m\ddot{x}_2 &= -2k x_2 + k x_1 + k x_3 \\
            m\ddot{x}_3 &= -2k x_3 + k x_2
        \end{cases}
    $$
    Assume solutions of the form:
    $$        
    \begin{cases}
            x_1(t) &= A \cos(\omega t + \phi) \\
            x_2(t) &= B \cos(\omega t + \phi) \\
            x_3(t) &= C \cos(\omega t + \phi)
        \end{cases}
    $$
    Substituting these assumed solutions into the equations of motion, we get:
    $$      
        \begin{cases}
            A \omega^2 &= -\frac{k}{m}(-2A + B) \\
            B \omega^2 &= -\frac{k}{m}(A - 2B + C) \\
            C \omega^2 &= -\frac{k}{m}(B - 2C)
        \end{cases}
    $$
    We can rewrite this as an eigenvalue problem:
    $$
        \begin{bmatrix}
            -\frac{2k}{m} & \frac{k}{m} & 0 \\
            \frac{k}{m} & -\frac{2k}{m} & \frac{k}{m} \\
            0 & \frac{k}{m} & -\frac{2k}{m}
        \end{bmatrix}
        \begin{bmatrix}
            A \\ B \\ C
        \end{bmatrix}
        = \omega^2
        \begin{bmatrix}
            A \\ B \\ C
        \end{bmatrix}
    $$
    To find the eigenvalues \( \omega^2 \), we solve the characteristic equation:
    $$
        \text{det}\left(\begin{bmatrix} 
            -\frac{2k}{m} - \omega^2 & \frac{k}{m} & 0 \\
            \frac{k}{m} & -\frac{2k}{m} - \omega^2 & \frac{k}{m} \\
            0 & \frac{k}{m} & -\frac{2k}{m} - \omega^2
        \end{bmatrix}\right) = 0
    $$
    Calculating the determinant, we have:
    $$        \left(\frac{2k}{m} - \omega^2\right)\left[\left(-\frac{2k}{m} - \omega^2\right)^2 - \left(\frac{k}{m}\right)^2\right] - \left(\frac{k}{m}\right)^2\left(\frac{2k}{m} - \omega^2\right) = 0
    $$   
    This gives us the three normal mode frequencies:
    $$
        \begin{cases}
            \omega_1 &= \sqrt{\frac{2k}{m}} \\
            \omega_2 &= \sqrt{\frac{2k}{m}(2 - \sqrt{2})} \\
            \omega_3 &= \sqrt{\frac{2k}{m}(2 + \sqrt{2})}
        \end{cases}
    $$
\end{example}
\paragraph{Intuition} For any perturbation, we can expressed it as a linear combination of the excitations of several normal modes. Each normal mode oscillates at its own frequency, and the overall motion is a superposition of these modes.
\section{Waves}

\subsection{Travelling Waves}

\begin{definition}[Travelling Pulse]
    A travelling pulse is a disturbance that moves through a medium. For any wave function $f(x,t)$, if it satisfies the property:
    \begin{equation} \label{eq:travelling_wave}
        y(x,t) = f(x \pm vt)
    \end{equation}
    then it represents a wave travelling in the positive (for $x - vt$) or negative (for $x + vt$) x-direction with speed \( v \). $y$ could represent displacement, pressure, electric field, etc.
\end{definition}

\begin{definition}[Sinusoidal Wave]
    A sinusoidal wave is a wave that can be described by a sine or cosine function. The general form of a sinusoidal wave travelling in the positive x-direction is:
    \begin{equation}
        y(x,t) = A \cos(kx \pm \omega t + \phi_0)
    \end{equation}
    where:
    \begin{itemize}
        \item \( A \) is the amplitude (maximum displacement)
        \item \( k = \frac{2\pi}{\lambda} \) is the wave number, with \( \lambda \) being the wavelength
        \item \( \omega = 2\pi f \) is the angular frequency, with \( f \) being the frequency
        \item \( \phi _0 \) is the initial phase.
    \end{itemize}
    So we can write the wave function as a form of \eqref{eq:travelling_wave}:
    \begin{equation}
        y(x,t) = A \cos\left[k(x \pm vt) + \phi_0\right]
    \end{equation}
    The wave speed \( v \) is related to the angular frequency and wave number by:
    \begin{equation}
        v = \frac{\omega}{k} = f\lambda
    \end{equation}
\end{definition}

\paragraph{Transverse and Longitudinal Waves} There is two (or three) main categories of waves:
\begin{itemize}
    \item \textbf{Transverse Waves}: In transverse waves, the oscillations are perpendicular to the direction of wave propagation. \\
    (Examples) Light waves, water waves, and waves on a string.
    \item \textbf{Longitudinal Waves}: In longitudinal waves, the oscillations are parallel to the direction of wave propagation. \\
    (Examples) Sound waves and pressure waves in fluids.
    \item \textbf{Both}: Some waves can exhibit both transverse and longitudinal characteristics. \\
    (Examples) Rayleigh surface waves in seismology.
\end{itemize}
\begin{definition}[Velocity of a Fixed Particle]
    For a fixed particle, we have:
    \begin{equation}
        v_y = \frac{\partial y}{\partial t} = A \omega \cos(kx \pm \omega t + \phi_0)
    \end{equation}
\end{definition}

\begin{example}[Vibrating String]
    Consider a string in the $xy$-plane, stretched along the x-axis with tension \( \tau \) and linear mass density \( \mu \). Now, consider at the end of the string at $x$ makes a small angle \( \theta \) with the x-axis. Then the components of the tension are:
    $$
    \begin{cases}
        \tau_x &= \tau \cos(\theta) \approx \tau \\
        \tau_y &= \tau \sin(\theta) \approx \tau \frac{\partial y}{\partial x}
    \end{cases}
    $$
    Now, consider a small segment of the string between \( x \) and \( x + \delta x \). The vetical force on the other end is given, via linear approximation, by:
    $$    
        \tau_y(x + \delta x) \approx \tau_y(x) \left[ \frac{\partial y}{\partial x}_x + \frac{\partial^2 y}{\partial x^2} \delta x \right]
    $$
    So the net force is given by:
    $$
        F_y = \tau_y(x + \delta x) - \tau_y(x) = \tau \frac{\partial^2 y}{\partial x^2} \delta x
    $$
    Becuase $F=ma$, we have:
    $$
        \tau \frac{\partial^2 y}{\partial x^2} \delta x = dm \frac{\partial^2 y}{\partial t^2}
    $$
    where \( dm = \mu \delta x \) is the mass of the small segment. Thus, we have:
    $$
        \frac{\partial^2 y}{\partial t^2} - v^2 \frac{\partial^2 y}{\partial x^2} = 0
    $$
    where \( v = \sqrt{\frac{\tau}{\mu}} \) is the wave speed on the string. This is the one-dimensional wave equation.

\subsection{The Wave Equation}
